---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Loading the requied packages
```{r, message = FALSE}
library("Hmisc", "lubridate")
```

unzipping and loading the data
```{r unzip activity.zip}
unzip("activity.zip")
activity <- read.csv("activity.csv")
```

viewing basic information about the dataset
```{r Explore}
str(activity)
head(activity, 5)
tail(activity, 5)
summary(activity)
```

Converting the date from class factor to class dates
```{r preprocess}
library("lubridate")
activity$date <- ymd(activity$date)
```

## What is mean total number of steps taken per day?
Calculating the total number of steps taken per day (ignoring missing values)
```{r Sum of steps per day}
steps_per_day <- aggregate(steps ~ date, activity, sum)
```

Plotting a Histogram of the total number of steps taken per day
```{r Hist}
ggplot(data = steps_per_day, mapping = aes(x = steps)) + 
        geom_histogram(fill = "limegreen", bins = 20) +
        labs(title = "Total number of steps taken each day", x = "steps per day", y = "Frequency")
```

Mean and median of the total number of steps taken per day
```{r mean and median}
mean(steps_per_day$steps, na.rm = TRUE)
median(steps_per_day$steps, na.rm = TRUE)
```

## What is the average daily activity pattern?
A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r timeseries1}
average_interval <- aggregate(steps ~ interval, activity, mean)
ggplot(data = average_interval, aes(x = interval, y = steps)) + 
        geom_line(color = "limegreen") +
        labs(title = "Average number of steps taken by 5-minute interval", 
             x = "time interval(mins)", y= "Average number of steps")
```

5-5-minute interval, on average across all the days in the dataset, containing the maximum number of steps
```{r maximum interval}
average_interval[which.max(average_interval$steps), 1]
```

## Imputing missing values
The total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r Number of NAs}
sum(is.na(activity$steps))
```

Filling in all of the missing values in the dataset
```{r Impute}
steps_imputed <- with(activity, impute(steps, mean))
```

Creating a new dataset that is equal to the original dataset but with the missing data filled in
```{r data with imputed steps}
activity2 <- cbind(steps_imputed, activity)
activity2 <- activity2[ , -2]
head(activity2, 10)
```

Plotting Histogram of the total number of steps taken each day 
```{r Hist_imputed}
daily_steps <- aggregate(steps_imputed ~ date, activity2, sum)

ggplot(data = daily_steps, mapping = aes(x = steps_imputed)) + 
        geom_histogram(fill = "limegreen", bins = 20) +
        labs(title = "Total number of steps taken each day", x = "steps per day", y = "Frequency")
```

Mean and median total number of steps taken per day
```{r Mean and median of imputed data}
Mean_steps <- mean(daily_steps$steps_imputed)
Median_steps <- median(daily_steps$steps_imputed)
```

## Are there differences in activity patterns between weekdays and weekends?
Creating a variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day
```{r weekdays and weekends}
day_type <- function(day){
        if(wday(day) %in% c(1,7))
                name <-"weekend"
        else
                name <-"weekday"
        name
}
activity2$day <- sapply(activity2$date, day_type)
head(activity2, 5)
```

A plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r, timeseries2}
average_interval2 <- aggregate(steps_imputed ~ interval + day, activity2, mean)

ggplot(data = average_interval2, aes(x = interval, y = steps_imputed)) + 
        geom_line(color = "limegreen") + 
        facet_wrap(day ~ .) +
        labs(title = "Average number of steps taken by time interval on weekdays and weekends",
            x = "time interval(mins)", y = "Average steps" )
```

